Guidance for clinicians in constructing ordinal outcomes and choosing ordinal effect measures for clinical trials

### 2025-05-15 notes_on_effect_measures.tex
Additional work

\begin{itemize}
  \item \citet{kruskalOrdinalMeasuresAssociation1958} even
    considers further involving \emph{three} independent copies of
    the variables.
  \item Covariances and least squares regression coefficients can also be
    expressed using the ``pairs of draws'' device:\footnote{ I believe this is
      connected with
      $U$-statistics.\url{https://stats.stackexchange.com/questions/612496/linear-regression-vs-average-of-slopes}
    } \[ \cov(x, y) = \frac{1}{2}\ave(x - x')(y - y') \\ b_{yx} =
      \frac{\ave (y -
      y')(x - x')}{\ave (x - x')^2} = {\ave \left(w \times
      \text{pair-slopes}\right)}
    \]
    ... where the weight $w_{ij}$ is ${\Delta x^2}\big/{\sum \Delta x^2}$.

@@

Then the ROC curve is a plot of $\Pr(y \leq t \mid X = \text{True})$
(which is just sensitivity)
versus $\Pr(y \leq t \mid X = \text{False})$ (which is just 1 - specificity).

@@

\subsubsection{Binary $x$, ordinal $y$}

\begin{itemize}
  \item \textbf{Mann-Whitney $U$ statistic and its limit} Rescaling
    gives estimate of ``common language effect size'' = probabilistic
    index = probability of superiority = AUROC = win probability.... See
    \url{https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test#Common_language_effect_size}.
      Note that this page also mentions a statistic ``$\rho$''
      which is the version incorporating ties, i.e.~$\Pr(Y > X) +
      0.5 \times \Pr(Y = X)$.
    \item \textbf{Wilcoxon statistic and its limit} Signed rank test. Either a
      single sample (in which case testing "whether the data come from
      a symmetric distribution with a specified center") or paired samples.
    \item \textbf{Win probability} Measure of effect. In terms of
      Somers' D it's
      $\frac{\text{SD} + 1}{2}$
    \item \textbf{Win difference} Seems to be the specialization of
      Kendall's Tau (see below).
    \item \textbf{Win ratio} Pr(win)/Pr(loss) (either matched or unmatched)
    \item \textbf{Win odds} Differs from the win ratio by incorporating ties
    \item \textbf{Mean ridit}
  \end{itemize}

  \subsubsection{Ordinal $x$, ordinal $y$}

@@

  \subsubsection{Ordinal $x$, ordinal $y$}

>>>>

\begin{itemize}
  \item \citet{kruskalOrdinalMeasuresAssociation1958} even
    considers further involving \emph{three} independent copies of
    the variables.
  \item Covariances and least squares regression coefficients can also be
    expressed using the ``pairs of draws'' device:\footnote{ I believe this is
      connected with
      $U$-statistics.\url{https://stats.stackexchange.com/questions/612496/linear-regression-vs-average-of-slopes}
      See also \url{https://www.fharrell.com/post/po/} discussion of
      Mann-Whitney. This also says that \textbf{it's possible to
        convert between Wilcoxon statistic, Somers' $D$, concordance
      probability, \emph{and} the Spearman rank correlation.}
    } \[ \cov(x, y) = \frac{1}{2}\ave(x - x')(y - y') \\ b_{yx} =
      \frac{\ave (y -
      y')(x - x')}{\ave (x - x')^2} = {\ave \left(w \times
      \text{pair-slopes}\right)}
    \]
    ... where the weight $w_{ij}$ is ${\Delta x^2}\big/{\sum \Delta x^2}$.

@@

Then the ROC curve is a plot of $\Pr(Y \leq t \mid X = \text{True})$
(which is just sensitivity)
versus $\Pr(Y \leq t \mid X = \text{False})$ (which is just 1 -
specificity) -- that is, a plot of cdf($y \mid x = 1$) versus cdf($y
\mid x = 0$)

@@

\subsubsection{Binary \emph{x}, ordinal \emph{y}}

\begin{itemize}
  \item \textbf{Mann-Whitney $U$ statistic and its limit} Rescaling
    gives estimate of ``common language effect size'' = probabilistic
    index = probability of superiority = AUROC = win probability.... See
    \url{https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test#Common_language_effect_size}.
      Note that this page also mentions a statistic ``$\rho$''
      which is the version incorporating ties, i.e.~$\Pr(Y > X) +
      0.5 \times \Pr(Y = X)$.
      % \item \textbf{Wilcoxon statistic and its limit} Signed rank
      % test. Either a
      %   single sample (in which case testing "whether the data come from
      %   a symmetric distribution with a specified center") or paired samples.
    \item \textbf{Win probability} Measure of effect. In terms of
      Somers' D it's
      $\frac{\text{SD} + 1}{2}$
    \item \textbf{Win difference} Seems to be the specialization of
      Kendall's Tau (see below).
    \item \textbf{Win ratio} Pr(win)/Pr(loss) (either matched or unmatched)
    \item \textbf{Win odds} Differs from the win ratio by incorporating ties
    \item \textbf{Mean ridit}
  \end{itemize}

  \subsubsection{Ordinal \emph{x}, ordinal \emph{y}}

@@

  \subsubsection{Ordinal \emph{x}, ordinal \emph{y}}

### 2025-05-15 paper.tex
Additional work

@@

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, e.g.
    comparative CDF plots.
  \item Ordinal regression models can use win methods for model checking; it is
    also possible to present a model-based win effect measure alongside a common
    odds ratio for example.
  \item The outcomes \emph{implicitly} constructed using
    DOOR and win methods may contain many levels, and may contain
    levels where an
    effect is not so meaningful (e.g. time-to-death when the followup time is
    short). By \emph{explicitly} constructing the associated ordinal outcome one
    might be forced to be more aware of this
  \item What is a reasonable win
    analysis of a specific component score? From the traditional
    ordinal endpoint
    school of thought what would make most sense is to \emph{only
      tie-break for the
    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
    variable. Is this what the plots do that break down wins and losses by
    component score?
\end{itemize}

>>>>

Also, we "chunk" longer hospital stays more aggressively to avoid
small categories, for interpretability's sake.

@@

Key references:
\citet{agrestiMeasuresNominalOrdinalAssociation1981,agrestiOrdinalProbabilityEffect2017a,
agrestiSimpleWaysInterpret2018a}

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, for the
    \emph{implicit} ordinal variable
    \begin{itemize}
      \item
        comparative CDF plots for the
      \item ridit plots
        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984}
        relative to a reference group, which should allow one to
        display both full eCDF information and also win statistics
        (based on the interpretation of the mean ridit as a win probability)
    \end{itemize}e.g.
  \item Ordinal regression models can use win methods for model checking; it is
    also possible to present a model-based win effect measure alongside a common
    odds ratio for example. See
    \citep{agrestiOrdinalProbabilityEffect2017} for example, which
    covers both latent variable and predicted probability versions.
  \item The outcomes \emph{implicitly} constructed using
    DOOR and win methods may contain many levels, and may contain
    levels where an
    effect is not so meaningful (e.g. time-to-death when the followup time is
    short). By \emph{explicitly} constructing the associated ordinal outcome one
    might be forced to be more aware of this
  \item What is a reasonable win
    analysis of a specific component score? From the traditional
    ordinal endpoint
    school of thought what would make most sense is to \emph{only
      tie-break for the
    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
    variable. Is this what the plots do that break down wins and losses by
    component score?
\end{itemize}
